---
title: "MODA • Step 01 — Import, ID Alignment & Common Samples"
author: "Abdulaziz"
date: "2025-09-14"
output: html_document
---

### ===============================================================

## Description:

### This R Markdown script executes Step 01 of the MODA pipeline.

### It imports and harmonizes multi-omics datasets, aligning all data layers

### (RNA, Proteome, Phospho, Methylation, miRNA, Acetylation) to a common

### set of valid samples for consistent downstream analyses.

## Key Functions:

### • Automatically detect and read raw omics data files.

### • Standardize feature and sample identifiers.

### • Identify common samples across all omics layers.

### • Save aligned matrices as both Parquet and RDS for efficiency.

### • Generate summary statistics and QC plots.

### • Perform acceptance checks to validate data alignment.

# 

## Outputs:

### • Processed omics data saved in: /data/processed/<layer>/

### • Common sample list: common_samples.rds

### • HTML report with summaries, plots, and QC metrics.

# 

## Summary:

### Provides a unified, analysis-ready dataset ensuring sample consistency

### across all omics modalities.

### ==============================================================

```{r}
# ===== 0) Package Setup =====
need <- c("tidyverse", "data.table", "janitor", "arrow", "ggpubr", "stringr")
inst <- need[!need %in% installed.packages()[, 1]]
if (length(inst)) install.packages(inst, repos = "https://cloud.r-project.org")


suppressPackageStartupMessages({
library(tidyverse)
library(data.table)
library(janitor)
library(arrow)
library(ggpubr)
library(stringr)
})

```

```{r}
# ===== 1) Project Paths =====
source("R/workflow_utils.R")
cfg <- load_moda_paths()
base_dir <- cfg$external_project_dir
raw_dir <- cfg$external_raw_dir
proc_dir <- cfg$external_processed_dir
if (!dir.exists(raw_dir)) stop("Raw omics directory not found. Update config/paths.yml.")
if (!nzchar(proc_dir)) stop("Processed data directory is not configured. Update config/paths.yml.")
dir.create(proc_dir, showWarnings = FALSE, recursive = TRUE)


# Helper functions
standardize_ids <- function(x) {
x <- toupper(x)
x <- gsub("\\s+", "", x) # Remove spaces
x <- gsub("[^A-Z0-9_-]", "", x) # Allow only A-Z, 0-9, _ and -
x
}


save_matrix_parquet <- function(m, out_path) {
df <- as.data.frame(m) %>% tibble::rownames_to_column("feature_id")
arrow::write_parquet(df, out_path)
}


# Flexible reader for *.cct.txt / *.txt files (detects first Feature/Gene column)
read_cct_matrix <- function(path) {
dt <- data.table::fread(path, check.names = FALSE)

# Guess feature column from known candidates
candidates <- c("GeneSymbol", "Gene", "SYMBOL", "Feature", "Protein", "ProteinID",
"miRNA", "miRNA_ID", "ID", "Accession")
rid <- if (length(intersect(candidates, names(dt)))) intersect(candidates, names(dt))[1] else names(dt)[1]
rn <- dt[[rid]]
drop_cols <- unique(c(rid, grep("(?i)desc|name|uniprot|protein.*name", names(dt), value = TRUE)))

# Remaining columns are samples
mat_df <- dt[, setdiff(names(dt), drop_cols), with = FALSE]

# Force numeric conversion
mat <- as.matrix(sapply(mat_df, as.numeric))
rownames(mat) <- rn
list(mat = mat, rid = rid, samples = colnames(mat))
}


```

```{r}
# ===== 2) Auto-detect file names from the raw directory =====
pick <- function(pattern)
  list.files(raw_dir, pattern = pattern, full.names = TRUE, ignore.case = TRUE)[1]

paths <- list(
  clinical = pick("^Clinical.*\\.txt$|\\.tsi\\.txt$"),
  rna      = pick("RNAseq.*TPM.*\\.cct(\\.txt)?$"),
  proteome = pick("proteome.*TumorNormal.*\\.cct(\\.txt)?$"),
  phospho  = pick("phospho.*geneLevel.*\\.cct(\\.txt)?$"),
  methyl   = pick("Methylation.*Gene.*\\.cct(\\.txt)?$"),
  mirna    = pick("microRNA.*RPM.*\\.cct(\\.txt)?$"),
  acetyl   = pick("acetylome.*geneLevel.*\\.cct(\\.txt)?$")
)

paths
```

```{r}
layers_to_read <- c("rna", "proteome", "phospho", "methyl", "mirna", "acetyl") # Main six layers

read_one <- function(p) {
  stopifnot(!is.na(p), file.exists(p))
  x <- read_cct_matrix(p)
  colnames(x$mat) <- standardize_ids(colnames(x$mat))
  x
}

objs <- lapply(paths[layers_to_read], read_one)
names(objs) <- layers_to_read

# Quick summary of loaded files
layer_summary <- tibble(
  layer          = names(objs),
  file           = unname(unlist(paths[layers_to_read])),
  feature_id_col = sapply(objs, `[[`, "rid"),
  n_features     = sapply(objs, function(z) nrow(z$mat)),
  n_samples      = sapply(objs, function(z) ncol(z$mat))
) %>% arrange(layer)

layer_summary %>% knitr::kable(caption = "Summary of layers after import")

```

```{r}
# Identify common samples across all layers
samples_per_layer <- lapply(objs, function(z) colnames(z$mat))
common_samples <- Reduce(intersect, samples_per_layer)

# Count number of common samples
common_n <- length(common_samples)
common_n
```

```{r}
# Plot the number of samples per layer
p_counts <- layer_summary %>%
  ggplot(aes(x = reorder(layer, n_samples), y = n_samples)) +
  geom_col() +
  geom_hline(yintercept = common_n, linetype = 2) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Number of Samples",
    title = "Number of Samples per Layer",
    subtitle = sprintf("Common samples across all layers = %d", common_n)
  )

p_counts

```

```{r}
# Display the first 20 common samples for quick review
tibble(sample_id = head(sort(common_samples), 20)) %>%
knitr::kable(caption = "First 20 common samples (quick review)")
```

```{r}
# Align all matrices to common samples
aligned <- lapply(objs, function(z) z$mat[, common_samples, drop = FALSE])

# Create directories for each omics layer
for (nm in names(aligned)) {
  dir.create(file.path(proc_dir, nm), showWarnings = FALSE, recursive = TRUE)
}

# Save matrices as Parquet and RDS (efficient and convenient)
mapply(function(m, nm) {
  out_parquet <- file.path(proc_dir, nm, paste0(nm, ".parquet"))
  out_rds     <- file.path(proc_dir, nm, paste0(nm, ".rds"))
  save_matrix_parquet(m, out_parquet)
  saveRDS(m, out_rds)
}, aligned, names(aligned))

# Save the list of common samples for later use
saveRDS(common_samples, file.path(proc_dir, "common_samples.rds"))

# Simple acceptance table
accept <- tibble(
  check  = c("Equal number of samples across matrices", "Number of common samples ≥ 250"),
  result = c(length(unique(sapply(aligned, ncol))) == 1, common_n >= 250),
  value  = c(unique(sapply(aligned, ncol))[1], common_n)
)

accept %>%
  knitr::kable(caption = "Acceptance Checks — Step 01")
```
