---
title: "MODA • Step 02 — QC & Normalization"
author: "Abdulaziz"
date: "2025-09-14"
output: html_document
---

# This R Markdown file performs systematic quality control (QC) and normalization

# across all omics layers processed in Step 01. It ensures clean, comparable,

# and batch-corrected datasets for downstream analyses.

# 

# Key Objectives:

## 1) Load processed omics matrices and clinical metadata.

## 2) Perform feature-level QC: remove excessive missingness, filter low variance.

## 3) Apply imputation (MinProb) and batch correction (ComBat) where appropriate.

## 4) Normalize data using Z-score scaling per feature.

## 5) Visualize QC through PCA plots before/after normalization.

## 6) Export cleaned and scaled datasets (RDS + Parquet) for reproducibility.

## 7) Summarize QC metrics: missingness, feature counts, and PCA separation (R²).

# 

# Outputs include:

## - PCA diagnostics for each omics layer.

## - Summary QC tables and feature statistics.

## - Harmonized, normalized omics datasets ready for integration.

# 

# This step establishes a unified foundation for robust multi-omics integration and modeling.

```{r}
# ===== 0) Lightweight package setup =====
need <- c("ggplot2", "data.table", "janitor", "Matrix", "matrixStats", "arrow", "sva")
inst <- need[!need %in% installed.packages()[, 1]]
if (length(inst)) install.packages(inst, repos = "https://cloud.r-project.org")


library(ggplot2)
library(data.table)
library(janitor)
library(Matrix)
library(matrixStats)
library(arrow)
library(sva)


# ===== 1) Define project paths =====
source("R/workflow_utils.R")
cfg <- load_moda_paths()
base_dir <- cfg$external_project_dir
proc_dir <- cfg$external_processed_dir
if (!nzchar(proc_dir) || !dir.exists(proc_dir)) {
  stop("Processed data directory not found. Update config/paths.yml.")
}


```

```{r}

# ===== 2) Utility Functions =====

# Z-score scaling (row-wise normalization)
zscale <- function(m) {
  m <- t(scale(t(as.matrix(m))))
  m[is.na(m)] <- 0
  m
}

# Calculate row-wise missingness ratio
row_miss <- function(m) rowMeans(is.na(m) | is.infinite(m))

# Filter features based on missingness threshold
keep_miss <- function(m, thr = 0.2) m[row_miss(m) <= thr, , drop = FALSE]

# PCA plotting function with optional metadata coloring
pca_plot <- function(m, meta = NULL, color_by = NULL, title = "PCA") {
  stopifnot(ncol(m) > 2)
  pc <- prcomp(t(m), center = TRUE, scale. = TRUE)
  df <- data.frame(sample_id = colnames(m), PC1 = pc$x[, 1], PC2 = pc$x[, 2])
  
  if (!is.null(meta) && !is.null(color_by) && color_by %in% names(meta)) {
    idx <- match(df$sample_id, meta$sample_id)
    df$col <- meta[[color_by]][idx]
    ggplot(df, aes(PC1, PC2, color = col)) +
      geom_point(size = 2) +
      labs(title = title, subtitle = paste("Color by:", color_by))
  } else {
    ggplot(df, aes(PC1, PC2)) +
      geom_point(size = 2) +
      labs(title = title)
  }
}

# Save processed and scaled versions of each omics layer
save_layer <- function(clean, scaled, layer) {
  out_dir <- file.path(proc_dir, layer)
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  saveRDS(clean,  file.path(out_dir, "clean.rds"))
  saveRDS(scaled, file.path(out_dir, "scaled.rds"))
  arrow::write_parquet(as.data.frame(clean) |> tibble::rownames_to_column("feature_id"),
                       file.path(out_dir, "clean.parquet"))
  arrow::write_parquet(as.data.frame(scaled) |> tibble::rownames_to_column("feature_id"),
                       file.path(out_dir, "scaled.parquet"))
}

# Standardize IDs: uppercase, remove spaces and non-alphanumeric characters
std_ids <- function(x) {
  x <- toupper(gsub("\\s+", "", as.character(x)))
  gsub("[^A-Z0-9_-]", "", x)
}

# Prepare data for PCA before normalization (filtering + median imputation)
prep_for_pca_before <- function(m, min_non_na_frac = 0.5) {
  m <- as.matrix(m)
  m[!is.finite(m)] <- NA
  
  # Require at least half the samples per feature (conservative threshold)
  min_ok <- pmax(3, ceiling(min_non_na_frac * ncol(m)))
  keep1 <- rowSums(!is.na(m)) >= min_ok
  m <- m[keep1, , drop = FALSE]
  
  # Impute missing values using median per feature
  if (anyNA(m)) {
    m <- t(apply(m, 1, function(x) { x[is.na(x)] <- median(x, na.rm = TRUE); x }))
  }
  
  # Remove features with zero variance
  keep2 <- matrixStats::rowSds(m) > 0
  m <- m[keep2, , drop = FALSE]
  m
}


```

```{r}
# ===== 3) Load Step 01 Matrices (directly from RDS) =====
layers <- c("rna", "proteome", "phospho", "methyl", "mirna", "acetyl")
mats <- setNames(lapply(layers, function(L) readRDS(file.path(proc_dir, L, paste0(L, ".rds")))), layers)
common_samples <- readRDS(file.path(proc_dir, "common_samples.rds"))
cs <- std_ids(common_samples)

# ===== 4) Load Clinical Metadata =====
clin_path <- list.files(file.path(base_dir, "data", "raw"),
                        pattern = "(?i)^Clinical.*\\.txt$|\\.tsi\\.txt$", full.names = TRUE)[1]

clinical <- data.table::fread(clin_path, data.table = FALSE)
names(clinical) <- janitor::make_clean_names(names(clinical))

# Select best column matching sample IDs
cands <- c("sample_id", "sample", "aliquot", "aliquot_id", "experimental_sample_id",
           "case_id", "subject_id", "tumor_sample_barcode")
present <- intersect(cands, names(clinical))
scores <- sapply(present, function(col) sum(std_ids(clinical[[col]]) %in% cs))
best_col <- present[which.max(scores)]
clinical$sample_id <- std_ids(clinical[[best_col]])

# Identify batch/type columns (manually defined)
batch_col <- if ("study_organization_0cptac_1icpc" %in% names(clinical)) "study_organization_0cptac_1icpc" else NA
type_col  <- if ("tissue_type_0nat_1tumor" %in% names(clinical))          "tissue_type_0nat_1tumor"          else NA

meta <- data.frame(
  sample_id = clinical$sample_id,
  batch = if (!is.na(batch_col)) clinical[[batch_col]] else NA,
  type  = if (!is.na(type_col))  clinical[[type_col]]  else NA,
  stringsAsFactors = FALSE
)

meta <- unique(meta[meta$sample_id %in% cs, , drop = FALSE])
meta$batch <- ifelse(meta$batch %in% c(0, "0"), "CPTAC",
               ifelse(meta$batch %in% c(1, "1"), "ICPC", as.character(meta$batch)))
meta$type  <- ifelse(meta$type  %in% c(0, "0"), "NAT",
               ifelse(meta$type  %in% c(1, "1"), "Tumor", as.character(meta$type)))

# ===== 5) Determine PCA color variable =====
color_var <- NULL
if ("type" %in% names(meta) && length(unique(na.omit(meta$type))) > 1) {
  color_var <- "type"
} else if ("batch" %in% names(meta) && length(unique(na.omit(meta$batch))) > 1) {
  color_var <- "batch"
}

# Display small metadata preview
head(meta, 10)


```

```{r}
# ===== 6) RNA Layer — QC, Filtering, and Batch Correction =====

rna <- mats$rna[, cs, drop = FALSE]
rna_before <- rna

# Filter out bottom 10% of features by variance
rv <- matrixStats::rowVars(rna, na.rm = TRUE)
cut <- as.numeric(quantile(rv, 0.10, na.rm = TRUE))
rna <- rna[rv > cut, , drop = FALSE]

# Apply ComBat batch correction if applicable
rna_clean <- rna
if (!is.null(color_var) && "batch" %in% names(meta) && length(unique(na.omit(meta$batch))) > 1) {
  b <- factor(meta$batch[match(colnames(rna), meta$sample_id)])
  mod <- model.matrix(~1, data = data.frame(b))
  rna_clean <- ComBat(as.matrix(rna), batch = b, mod = mod, par.prior = TRUE, prior.plots = FALSE)
}

# Z-score scaling
rna_scaled <- zscale(rna_clean)

# PCA plots before and after batch correction
rna_before_for_pca <- prep_for_pca_before(rna_before)
p1 <- pca_plot(rna_before_for_pca, meta, color_by = color_var, title = "RNA • PCA (Before)")
p2 <- pca_plot(rna_scaled, meta, color_by = color_var, title = "RNA • PCA (After)")

p1; p2

# Save processed and scaled data + summary table
save_layer(rna_clean, rna_scaled, "rna")
data.frame(
  layer = "RNA",
  features_before = nrow(rna_before),
  features_after  = nrow(rna_scaled),
  samples         = ncol(rna_scaled)
)

```

```{r}
# ===== 7) Proteome Layer — QC, Imputation, and Batch Correction =====

prot <- mats$proteome[, cs, drop = FALSE]
prot_before <- prot

# Remove features with >20% missingness
prot <- keep_miss(prot, 0.2)

# Impute missing values using MinProb (if applicable)
prot_imp <- if (any(is.na(prot))) imputeLCMD::impute.MinProb(as.matrix(prot), q = 0.01) else prot

# Apply ComBat batch correction if batch information is available
prot_clean <- prot_imp
if (!is.null(color_var) && "batch" %in% names(meta) && length(unique(na.omit(meta$batch))) > 1) {
  b <- factor(meta$batch[match(colnames(prot_imp), meta$sample_id)])
  mod <- model.matrix(~1, data = data.frame(b))
  prot_clean <- ComBat(as.matrix(prot_imp), batch = b, mod = mod, par.prior = TRUE, prior.plots = FALSE)
}

# Z-score scaling
prot_scaled <- zscale(prot_clean)

# PCA plots before and after normalization
prot_before_for_pca <- prep_for_pca_before(prot_before)
p1 <- pca_plot(prot_before_for_pca, meta, color_by = color_var, title = "Proteome • PCA (Before)")
p2 <- pca_plot(prot_scaled, meta, color_by = color_var, title = "Proteome • PCA (After)")
p1; p2

# Save processed and scaled data + summary table
save_layer(prot_clean, prot_scaled, "proteome")

data.frame(
  layer = "Proteome",
  features_before = nrow(prot_before),
  features_after  = nrow(prot_scaled),
  samples         = ncol(prot_scaled),
  mean_missing_after = mean(is.na(prot_imp))
)


```

```{r}
# ===== 8) Phospho Layer — QC, Imputation, and Batch Correction =====

library(imputeLCMD)

ph <- mats$phospho[, cs, drop = FALSE]
ph_before <- ph

# Remove features with >50% missingness
ph <- keep_miss(ph, 0.5)

# Impute missing values using MinProb (if applicable)
ph_imp <- if (any(is.na(ph))) imputeLCMD::impute.MinProb(as.matrix(ph), q = 0.01) else ph

# Apply ComBat batch correction if batch information is available
ph_clean <- ph_imp
if (!is.null(color_var) && "batch" %in% names(meta) && length(unique(na.omit(meta$batch))) > 1) {
  b <- factor(meta$batch[match(colnames(ph_imp), meta$sample_id)])
  mod <- model.matrix(~1, data = data.frame(b))
  ph_clean <- ComBat(as.matrix(ph_imp), batch = b, mod = mod, par.prior = TRUE, prior.plots = FALSE)
}

# Z-score scaling
ph_scaled <- zscale(ph_clean)

# PCA plot after normalization
pca_plot(ph_scaled, meta, color_by = color_var, title = "Phospho • PCA (After)")

# Save processed and scaled data + summary table
save_layer(ph_clean, ph_scaled, "phospho")

data.frame(
  layer = "Phospho",
  features_before = nrow(ph_before),
  features_after  = nrow(ph_scaled),
  samples         = ncol(ph_scaled),
  mean_missing_after = mean(is.na(ph_imp))
)

```

```{r}
# ===== 9) Methylation Layer — Logit Transform and Scaling =====

met <- mats$methyl[, cs, drop = FALSE]
met_before <- met

# Apply logit transform if data is within [0,1]
rng <- range(met, na.rm = TRUE)
met_m <- if (rng[1] >= 0 && rng[2] <= 1) log2((met + 1e-6) / (1 - met + 1e-6)) else met

# Z-score scaling
met_scaled <- zscale(met_m)

# PCA plot after normalization
pca_plot(met_scaled, meta, color_by = color_var, title = "Methylation • PCA (After)")

# Save processed and scaled data + summary table
save_layer(met_m, met_scaled, "methyl")

data.frame(
  layer = "Methylation",
  value_range_before = paste(round(rng, 3), collapse = " to "),
  features_after = nrow(met_scaled),
  samples = ncol(met_scaled)
)
```

# 6) miRNA — فلترة قليل التباين + Scaling

```{r}
# ===== 10) miRNA Layer — Variance Filtering and Scaling =====

mi <- mats$mirna[, cs, drop = FALSE]
mi_before <- mi

# Filter out bottom 10% of features by variance
rv <- matrixStats::rowVars(mi, na.rm = TRUE)
cut <- as.numeric(quantile(rv, 0.10, na.rm = TRUE))
mi_filt <- mi[rv > cut, , drop = FALSE]

# Z-score scaling
mi_scaled <- zscale(mi_filt)

# PCA plot after normalization
pca_plot(mi_scaled, meta, color_by = color_var, title = "miRNA • PCA (After)")

# Save processed and scaled data + summary table
save_layer(mi_filt, mi_scaled, "mirna")

data.frame(
  layer = "miRNA",
  features_before = nrow(mi_before),
  features_after  = nrow(mi_scaled),
  samples         = ncol(mi_scaled)
)

```

```{r}
# ===== 11) Acetyl Layer — QC, Imputation, and Batch Correction =====

ac <- mats$acetyl[, cs, drop = FALSE]
ac_before <- ac

# Remove features with >50% missingness
ac <- keep_miss(ac, 0.5)

# Impute missing values using MinProb (if applicable)
ac_imp <- if (any(is.na(ac))) imputeLCMD::impute.MinProb(as.matrix(ac), q = 0.01) else ac

# Apply ComBat batch correction if batch information is available
ac_clean <- ac_imp
if (!is.null(color_var) && "batch" %in% names(meta) && length(unique(na.omit(meta$batch))) > 1) {
  b <- factor(meta$batch[match(colnames(ac_imp), meta$sample_id)])
  mod <- model.matrix(~1, data = data.frame(b))
  ac_clean <- ComBat(as.matrix(ac_imp), batch = b, mod = mod, par.prior = TRUE, prior.plots = FALSE)
}

# Z-score scaling
ac_scaled <- zscale(ac_clean)

# PCA plot after normalization
pca_plot(ac_scaled, meta, color_by = color_var, title = "Acetyl • PCA (After)")

# Save processed and scaled data + summary table
save_layer(ac_clean, ac_scaled, "acetyl")

data.frame(
  layer = "Acetyl",
  features_before = nrow(ac_before),
  features_after  = nrow(ac_scaled),
  samples         = ncol(ac_scaled),
  mean_missing_after = mean(is.na(ac_imp))
)


```

# 8) Summary Panel (جاهز للعرض/الحفظ)

```{r}
# ===== 12) Summary of Features After QC and Normalization =====

summ <- rbind(
  data.frame(layer = "RNA",      features = nrow(readRDS(file.path(proc_dir, "rna", "clean.rds")))),
  data.frame(layer = "Proteome", features = nrow(readRDS(file.path(proc_dir, "proteome", "clean.rds")))),
  data.frame(layer = "Phospho",  features = nrow(readRDS(file.path(proc_dir, "phospho", "clean.rds")))),
  data.frame(layer = "Methyl",   features = nrow(readRDS(file.path(proc_dir, "methyl", "clean.rds")))),
  data.frame(layer = "miRNA",    features = nrow(readRDS(file.path(proc_dir, "mirna", "clean.rds")))),
  data.frame(layer = "Acetyl",   features = nrow(readRDS(file.path(proc_dir, "acetyl", "clean.rds"))))
)

# Display summary table
knitr::kable(summ, caption = "Number of features after cleaning for each layer")

# Bar plot of cleaned feature counts
ggplot(summ, aes(reorder(layer, features), features)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Features (clean)", title = "Features after QC/Normalization")
```

```{r}
# ===== 13) Missingness Summary Before Cleaning =====

# Combine raw matrices before filtering/imputation
mats_before <- list(
  RNA         = rna_before,
  Proteome    = prot_before,
  Phospho     = ph_before,
  Methylation = met_before,
  miRNA       = mi_before,
  Acetyl      = ac_before
)

# Compute missing proportion for each layer
miss_layer <- do.call(rbind, lapply(names(mats_before), function(nm) {
  m <- mats_before[[nm]]
  data.frame(layer = nm, missing_prop = mean(is.na(m) | !is.finite(m)))
}))

# Display summary table
knitr::kable(miss_layer, caption = "Overall missing value proportion per omics layer (before cleaning)")

# Visualize missingness
ggplot2::ggplot(miss_layer, ggplot2::aes(x = reorder(layer, -missing_prop), y = missing_prop)) +
  ggplot2::geom_col() +
  ggplot2::ylim(0, 1) +
  ggplot2::labs(x = NULL, y = "Missing proportion", title = "Overall Missingness per Omics (Before)")

```
