---
title: "MODA - Step 06 - DIABLO Supervised Integration"
author: "Abdulaziz"
date: "2025-09-27"
output: html_document
---


```{r setup}
suppressPackageStartupMessages({
library(mixOmics)
library(data.table)
library(ggplot2)
library(pROC)
})
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}

# ==== 1) Paths ====
source("R/workflow_utils.R")
cfg <- load_moda_paths()
base_dir <- cfg$external_project_dir
proc_dir <- cfg$external_processed_dir
results_root <- if (nzchar(cfg$external_results_dir)) cfg$external_results_dir else file.path(base_dir, "results")
res_dir  <- file.path(results_root, "diablo"); dir.create(res_dir, recursive=TRUE, showWarnings=FALSE)

```



```{r}
# ==== 2) Load scaled matrices ====
rna  <- readRDS(file.path(proc_dir,"rna","scaled.rds"))
prot <- readRDS(file.path(proc_dir,"proteome","scaled.rds"))
ph   <- readRDS(file.path(proc_dir,"phospho","scaled.rds"))
meth <- readRDS(file.path(proc_dir,"methyl","scaled.rds"))
mir  <- readRDS(file.path(proc_dir,"mirna","scaled.rds"))
ac   <- readRDS(file.path(proc_dir,"acetyl","scaled.rds"))

# Fix omics matrices (transpose)
fix_matrix <- function(mat){
  mat <- t(mat)                          # transpose
  mat <- as.data.frame(mat)              # return data.frame
  # make colnames = features, rownames = samples
  return(mat)
}

rna  <- fix_matrix(rna)
mir  <- fix_matrix(mir)
prot <- fix_matrix(prot)
ph   <- fix_matrix(ph)
meth <- fix_matrix(meth)
ac   <- fix_matrix(ac)
```


```{r}
# ==== 3) Load metadata ====
meta_orig <- as.data.frame(readRDS(file.path(proc_dir, "clinical", "meta_labels.rds")))
sample_col <- if ("sample" %in% names(meta_orig)) "sample" else "sample_id"
meta <- meta_orig
rownames(meta) <- meta[[sample_col]]
if (!"sample" %in% names(meta)) meta$sample <- meta[[sample_col]]

# outcome column (e.g., EGFR mutation status)
target_candidates <- c("egfr_mutation_info", "egfr_any_vs_wt", "egfr_classic_vs_other")
target_col <- target_candidates[target_candidates %in% names(meta)][1]
if (is.na(target_col)) {
  stop("No EGFR target column found in metadata. Expected one of: ", paste(target_candidates, collapse = ", "))
}
Y <- factor(meta[[target_col]])
```


```{r}
# ---- 3) Ensure sample names across blocks match ----
common_samples <- Reduce(intersect, list(
  rownames(rna), rownames(mir), rownames(prot),
  rownames(ph), rownames(meth), rownames(ac),
  rownames(meta)
))

length(common_samples)  # should be > 0


```

```{r}

# ---- 4) Subset and align row order ----
rna  <- rna[common_samples, ]
mir  <- mir[common_samples, ]
prot <- prot[common_samples, ]
ph   <- ph[common_samples, ]
meth <- meth[common_samples, ]
ac   <- ac[common_samples, ]
meta <- meta[common_samples, ]

stopifnot(all(rownames(meta) == rownames(rna)))

```


```{r}

# ==== 5) List of blocks ====
data.list <- list(
  RNA     = rna,
  miRNA   = mir,
  Methyl  = meth,
  Protein = prot,
  Phospho = ph,
  Acetyl  = ac
)

# ==== 6) Design matrix ====
design <- matrix(0.1, ncol = length(data.list), nrow = length(data.list))
diag(design) <- 0
colnames(design) <- rownames(design) <- names(data.list)

```



```{r}
# ==== 7) Tune keepX ====

set.seed(123)

tune_diablo <- tune.block.splsda(
  X = data.list,
  Y = Y,
  ncomp = 3,   # number of components
 test.keepX = list(
    RNA     = c(50,150),
    miRNA   = c(10,30),
    Methyl  = c(50,100),
    Protein = c(10,20),
    Phospho = c(10,20),
    Acetyl  = c(10,20)
  ),
  design = design,
  validation = "Mfold",
  folds = 5,
  nrepeat = 5,
  dist = "centroids.dist",
  progressBar = TRUE
)
```



```{r}
saveRDS(tune_diablo, file.path(base_dir, "results/diablo/diablo_tune_colab.rds"))
tune_diablo <- readRDS(file.path(base_dir, "results/diablo/diablo_tune_colab.rds"))

```



```{r}
# ==== 6) Get best keepX ====
best_keepX <- tune_diablo$choice.keepX
print(best_keepX)
```


```{r}
# ==== 7) Build final model with tuned keepX ====
diablo_model <- block.splsda(
  X = data.list,
  Y = Y,
  ncomp = 3,
  design = design,
  keepX = best_keepX
)
saveRDS(diablo_model, file.path(res_dir, "diablo_model_final.rds"))

```


```{r}
perf_diablo <- perf(diablo_model, validation = "Mfold",
                    folds = 5, nrepeat = 10, dist = "centroids.dist")
saveRDS(perf_diablo, file.path(res_dir, "diablo_perf_final.rds"))

```

```{r}
# Error rate per component
pdf(file.path(res_dir, "perf_diablo.pdf"), 
    width = 12,  
    height = 10) 
plot(perf_diablo)


# Finish plotting and save the file
dev.off()
```

```{r}
perf_diablo$choice.ncomp$WeightedVote

```



```{r}
plotDiablo(diablo_model, ncomp = 1)

pdf(file.path(res_dir, "plotDiablo1.pdf"), 
    width = 12,  
    height = 10) 
plotDiablo(diablo_model, ncomp = 1)


# Finish plotting and save the file
dev.off()

pdf(file.path(res_dir, "plotDiablo2.pdf"), 
    width = 12,  
    height = 10) 
plotDiablo(diablo_model, ncomp =2)


# Finish plotting and save the file
dev.off()

pdf(file.path(res_dir, "plotDiablo3.pdf"), 
    width = 12,  
    height = 10) 
plotDiablo(diablo_model, ncomp = 3)


# Finish plotting and save the file
dev.off()

```

```{r}
plotIndiv("diablo_model")

```


```{r}
# ==== 9) Extract and export selected features per block ====

extract_block_features <- function(model, block_name, res_dir) {
  sel <- selectVar(model, block = block_name)
  
  if (is.null(sel) || length(sel) == 0) {
    message("⚠️ No features selected for block: ", block_name)
    return(NULL)
  }
  
  out_list <- list()
  
  for (comp in names(sel)) {
    comp_df <- as.data.frame(sel[[comp]])
    
    # Auto-detect columns
    cols <- tolower(names(comp_df))
    if ("name" %in% cols) {
      feat_col <- "name"
    } else if ("names" %in% cols) {
      feat_col <- "names"
    } else {
      feat_col <- names(comp_df)[1]  # fallback to first column
    }
    
    if ("value" %in% cols) {
      val_col <- "value"
    } else if ("loading" %in% cols) {
      val_col <- "loading"
    } else if (ncol(comp_df) > 1) {
      val_col <- names(comp_df)[2]
    } else {
      val_col <- NA
    }
    
    comp_df$feature   <- comp_df[[feat_col]]
    comp_df$component <- comp
    comp_df$weight    <- if (!is.na(val_col)) comp_df[[val_col]] else NA_real_
    
    out_list[[comp]] <- comp_df[, c("feature", "weight", "component")]
  }
  
  df <- data.table::rbindlist(out_list, use.names = TRUE, fill = TRUE)
  data.table::fwrite(df, file.path(res_dir, paste0("DIABLO_sel_", block_name, ".csv")))
  
  message("✅ Saved: ", paste0("DIABLO_sel_", block_name, ".csv"),
          " (", nrow(df), " features)")
  return(df)
}

# Extract for each block
sel_rna   <- extract_block_features(diablo_model, "RNA", res_dir)
sel_mirna <- extract_block_features(diablo_model, "miRNA", res_dir)
sel_meth  <- extract_block_features(diablo_model, "Methyl", res_dir)
sel_prot  <- extract_block_features(diablo_model, "Protein", res_dir)
sel_ph    <- extract_block_features(diablo_model, "Phospho", res_dir)
sel_ac    <- extract_block_features(diablo_model, "Acetyl", res_dir)


```


```{r}
plotIndiv(
  diablo_model,
  comp = c(1, 2),
  group = diablo_model$Y,
  legend = TRUE,
  title = "DIABLO: Individuals (Samples) — Comp 1 vs 2" ,
ind.names = FALSE
)

pdf(file.path(res_dir, "plotIndiv 1_3.pdf"), 
    width = 12,  
    height = 10) 
plotIndiv(
  diablo_model,
  comp = c(1, 3),
  group = diablo_model$Y,
  legend = TRUE,
  title = "DIABLO: Individuals (Samples) — Comp 1 vs 3" ,
ind.names = FALSE
)
# Finish plotting and save the file
dev.off()

```


```{r}
plotVar(
  diablo_model,
  comp = c(1, 2),
  var.names = F,
  legend = TRUE,
  style = "graphics",
  title = "DIABLO: Correlation circle plot"
)

pdf(file.path(res_dir, "plotVar 1_3.pdf"), 
    width = 12,  
    height = 10) 

plotVar(
  diablo_model,
  comp = c(1, 3),
  var.names = F,
  legend = TRUE,
  style = "graphics",
  title = "DIABLO: Correlation circle plot"
)
# Finish plotting and save the file
dev.off()

```


```{r}
plotLoadings(
  diablo_model,
  comp = 1,
  contrib = "max",
  method = "mean",
  ndisplay = 20,
  title = "Top loadings (Component 1)"
)


pdf(file.path(res_dir, "plotLoadings_3.pdf"), 
    width = 12,  
    height = 10) 

plotLoadings(
  diablo_model,
  comp = 3,
  contrib = "max",
  method = "mean",
  ndisplay = 20,
  title = "Top loadings (Component 1)"
)
# Finish plotting and save the file
dev.off()

```

```{r}
plotArrow(diablo_model, ind.names = FALSE, legend = TRUE, 
          title = 'DIABLO comp 1 - 2')
pdf(file.path(res_dir, "plotArrow.pdf"), 
    width = 12,  
    height = 10) 

plotArrow(diablo_model, ind.names = FALSE, legend = TRUE, 
          title = 'DIABLO comp 1 - 2')
# Finish plotting and save the file
dev.off()
```

```{r}
circosPlot(diablo_model, cutoff = 0.7, line = TRUE, 
           color.blocks = c('darkorchid', 'brown1', 'lightgreen','blue','yellow','orange'),
           color.cor = c("chocolate3","grey20"), size.labels = 1.5)

pdf(file.path(res_dir, "circosPlot9.pdf"), 
    width = 12,  
    height = 10) 
circosPlot(diablo_model, cutoff = 0.9, line = TRUE, 
           color.blocks = c('darkorchid', 'brown1', 'lightgreen','blue','yellow','orange'),
           color.cor = c("chocolate3","grey20"), size.labels = 1.5)

# Finish plotting and save the file
dev.off()
```


```{r}
png("cimDiablo_plot.png")
cimDiablo(diablo_model, color.blocks = c('darkorchid', 'brown1', 'lightgreen','blue','yellow','orange'),
          comp = 1, margin = c(8, 20), legend.position = "right")

pdf(file.path(res_dir, "cimDiablo3.pdf"), 
    width = 12,  
    height = 10) 

cimDiablo(diablo_model, color.blocks = c('darkorchid', 'brown1', 'lightgreen','blue','yellow','orange'),
          comp = 3, margin = c(8, 20), legend.position = "right")
# Finish plotting and save the file
dev.off()
```

```{r}
network(
  diablo_model,
  comp = 1,
  threshold = 0.7,
  title = "DIABLO Network (comp1, corr > 0.7)"
)
pdf(file.path(res_dir, "network1.pdf"), 
    width = 12,  
    height = 10) 

network(
  diablo_model,
  comp = 1,
  threshold = 0.7,
  title = "DIABLO Network (comp1, corr > 0.7)"
)
# Finish plotting and save the file
dev.off()
```


```{r}
perf_diablo$MajorityVote.error.rate

```


```{r}
perf_diablo$WeightedVote.error.rate

```


```{r}

pdf(file.path(res_dir, "auroc3_Proteome.pdf"), 
    width = 12,  
    height = 10) 

auroc(diablo_model, roc.block = "Proteome", roc.comp =3,
                   print = FALSE)
# Finish plotting and save the file
dev.off()

pdf(file.path(res_dir, "auroc3_Phospho.pdf"), 
    width = 12,  
    height = 10) 

auroc(diablo_model, roc.block = "Phospho", roc.comp =3,
                   print = FALSE)
# Finish plotting and save the file
dev.off()

pdf(file.path(res_dir, "auroc3_Methyl.pdf"), 
    width = 12,  
    height = 10) 

auroc(diablo_model, roc.block = "Methyl", roc.comp =3,
                   print = FALSE)
# Finish plotting and save the file
dev.off()

pdf(file.path(res_dir, "auroc3_miRNA.pdf"), 
    width = 12,  
    height = 10) 

auroc(diablo_model, roc.block = "miRNA", roc.comp =3,
                   print = FALSE)
# Finish plotting and save the file
dev.off()

pdf(file.path(res_dir, "auroc3_Acetyl.pdf"), 
    width = 12,  
    height = 10) 

auroc(diablo_model, roc.block = "Acetyl", roc.comp =3,
                   print = FALSE)
# Finish plotting and save the file
dev.off()


```


```{r}
library(ggplot2)
library(pROC)

# your blocks
blocks <- c("RNA", "Proteome", "Phospho", "Methyl", "miRNA", "Acetyl")

# compute ROC for each block
roc_list <- list()
for (blk in blocks) {
  res <- auroc(diablo_model, roc.block = blk, roc.comp = 3, print = FALSE)
  # extract the first component (adjust if needed)
  roc_obj <- res[[blk]][[1]]
  roc_list[[blk]] <- roc_obj
}

# bind into a single DataFrame
roc_data <- do.call(rbind, lapply(names(roc_list), function(blk) {
  roc_obj <- roc_list[[blk]]
  data.frame(
    FPR = 1 - roc_obj$specificities,
    TPR = roc_obj$sensitivities,
    Block = blk,
    AUC = round(as.numeric(roc_obj$auc), 3)
  )
}))

# plot using ggplot2
p <- ggplot(roc_data, aes(x = FPR, y = TPR, color = Block)) +
  geom_line(size = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "ROC Curves for All Omics (Comp 3)",
    x = "1 - Specificity (%)",
    y = "Sensitivity (%)",
    color = "Omic Block"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# add AUC values into legend
auc_labels <- sapply(split(roc_data, roc_data$Block), function(df) unique(df$AUC))
auc_labels <- paste(names(auc_labels), ": AUC =", auc_labels)
p <- p + scale_color_discrete(name = "Outcome", labels = auc_labels)

# save plot
ggsave(file.path(res_dir, "auroc_all_omics_combined.pdf"), p, width = 10, height = 8)
